Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 $M_\textrm{w}$ 7.9 Wenchuan earthquake

### 2014-10-27 paper.tex
BJM edits to cover letter and a bit more

\section{Abstract}
The Longmen Shan is the steepest topographic front of the India-Asia collision and was the site of the $\textrm{M}_{\textrm{w}}$ 7.9  Wenchuan earthquake. Shortening estimates across the Longmen Shan provide strain accumulation rates and clarify the eastward extrusion of the Tibetan plateau. Here, to explain the interseismic GPS velocities across the greater Longmen Shan region, we develop a boundary element model including earthquake cycle effects, topography, the westward dipping Beichuan fault, and a ${\sim}20$ km deep, shallowly dipping, detachment. The detachment is inferred from observations of slip during and after the Wenchuan earthquake and from structural considerations. Previous analyses which neglected the detachment and earthquake cycle effects have found shortening rates near zero. In contrast, we find that interseismic GPS data are consistent with a shortening rate of 5.7$\pm$1.5 mm/yr and maximum surface slip-deficit rate of 9.5 $\pm$ 2.5 mm/yr. This model provides a unified interpretation of geodetic deformation throughout the earthquake cycle and suggests that the Longmen Shan is an active fold-and-thrust belt with Wenchuan style earthquake recurrence intervals of as short as 600 years.

>>>>

\section{Cover Letter}
Dear Geology editors,

Please find attached a manuscript entitled ``Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 $M_\textrm{w}$ 7.9 Wenchuan earthquake'' by T. Ben Thompson and Brendan J. Meade. We are submitting this manuscript for consideration as an article in Geology. The manuscript contains 3249 words and 4 figures and we are happy to pay for color figures.

The 2008 $M_\textrm{w}$ 7.9 Wenchuan earthquake led to the loss of more than 60,000 lives and prompted a fundamental reconsideration of how we understand the active tectonics of the remarkably steep Longmen Shan rangefront at the eastern margin of the Tibet Plateau. In particular, this seismic event has brought into sharp relief the debate an understanding of the rangefront as a active fold-and-thrust belt or as the passive response to the deeper flow of lower crustal material.

Geodetic slip and slip-deficit rates are a primary way of quantifying fault system kinematics and providing constraints on earthquake recurrence intervals. Previous estimates of slip rates that the Longmen Shan have been exceptionally crude (and serve only as a lower bound) due to the neglect of earthquake cycle effects and . Recent observations of co- and post-seismic slip on a deep detachment suggest that Longmen Shan regional fault geometry may be more complex than previously thought. Using a new boundary element model including the effects of both topography and non-planar fault system geometry, we analyze the implications of a interseismically locked detachment beneath eastern Tibet and show that geometry can produce a velocity gradient that is steepest 100-200 km northwest of the Longmen Shan rangefront, exactly as geodetically observed, with a negligible velocity gradient at the steepest part of the rangefront. Interpreting GPS velocities measured prior to the Wenchuan earthquake, inferred slip-deficit rates on the Beichuan fault may be as high as 10 mm/yr at the surface. These results show a simple analysis of the GPS velocities across the Longmen Shan will produce dramatically different results than an analysis including more detailed fault geometry.

These results are of wide significance for three reasons: 1) Providing a new interpretation of tectonics and earthquake cycle physics at the Longmen Shan, 2) Demonstrating how topography can be explicitly incorporated into earthquake cycle problems, and 3) Illustrating the complex role the deep detachment faults can play in controlling earthquake cycle behaviors. We are optimistic that these results will not only contribute significantly to the debate about the tectonics of the Tibetan Plateau and, more generally, about the nature of the development of fold-and-thrust belts.  Geology has served as the journal for the greatest insights on the  active tectonics and seismic hazard of eastern Tibet (King et al,. 1997; Clark and Royden, 2000; Zhang et al., 2004; Clark et al., 2005; Meade, 2007; Xu et al., 2009; Zhao et al., 2012; Wang et al., 2014) and our work offers a substantial conceptual and quantitative revisions of both previous tectonic concepts and earthquake recurrence intervals at this most active of margins.

Thanks for your thoughtful consideration of this paper. We're looking forward to hearing from you soon!

T. Ben Thompson

Department of Earth \& Planetary Sciences, Harvard University

Cambridge, MA 02138

(617)-NNN-NNNN, tthompson@fas.harvard.edu

\section{Abstract}
The Longmen Shan is the steepest topographic front assocated with the India-Asia collision and was the site of the $\textrm{M}_{\textrm{w}}$ 7.9  Wenchuan earthquake. Shortening estimates across the Longmen Shan provide strain accumulation rates and clarify the eastward extrusion of the Tibetan plateau. Here, to explain the interseismic GPS velocities across the greater Longmen Shan region, we develop a boundary element model including earthquake cycle effects, topography, the westward dipping Beichuan fault, and a ${\sim}20$ km deep, shallowly dipping, detachment. The detachment is inferred from observations of slip during and after the Wenchuan earthquake and from structural considerations. Previous analyses which neglected the detachment and earthquake cycle effects have found shortening rates near zero. In contrast, we find that interseismic GPS data are consistent with a shortening rate of 5.7$\pm$1.5 mm/yr and maximum surface slip-deficit rate of 9.5 $\pm$ 2.5 mm/yr. This model provides a unified interpretation of geodetic deformation throughout the earthquake cycle and suggests that the Longmen Shan is an active fold-and-thrust belt with Wenchuan style earthquake recurrence intervals of as short as 600 years.
